---
title: "Assignment 1-3"
author: "YOUR NAME(S) HERE"
date: "February 17, 2019"
output:
  html_document: default
  pdf_document: default
---

##### **Part 1: Data visualization**
Improve or refine your data visualization to address question Q1.1: "How does the risk of disease compare for smokers and otherwise similar non-smokers?"

As before, you'll need to start by loading in the data set -- you can copy and paste the path you used in your previous assignments.
```{r}
nmes.data<-read.csv("nmesUNPROC.csv", stringsAsFactors=FALSE)
```



```{r}

```


##### **Part 2: Logistic regression modeling**

Based on what we discussed in class, update your logistic regression model to address question Q1.1 and interpret the results. What does the model say about Q1.1? *Be sure to focus on answering the question being asked.*

```{r}

```

##### **Part 3: Propensity score calculation**

As discussed in class, use logistic regression to estimate propensity scores for the treatment of smoking (eversmk); that is, use logistic regression to estimate the probability of smoking given possible confounders.

```{r}

```

Then, using these propensity scores, use logistic regression with quintiles of your propensity scores to answer Q1.1. Interpret the results.

```{r}

```
