---
title: "Class 1-3"
author: "Margaret Taub and Leah Jager"
date: "February 11, 2019"
output:
  html_document: default
  pdf_document: default
---

This file contains some programming topics that are relevant to your assignment for this week.

##### **Part 1: Calculating and working with propensity scores**

Since you will be using propensity scores in your assignment this week, here is an example of how to estimate them in `R`.

First, we need to load our data set and recode some variables that we will use.
```{r}
nmes.data<-read.csv("nmesUNPROC.csv", stringsAsFactors=FALSE)
nmes.data$female<-factor(nmes.data$female, levels=c(0,1), labels=c("Male", "Female"))
nmes.data$eversmk<-factor(nmes.data$eversmk, levels=c(0,1), labels=c("No", "Yes"))
nmes.data$lc5<-factor(nmes.data$lc5, levels=c(0,1), labels=c("No LC", "LC"))
nmes.data$chd5<-factor(nmes.data$chd5, levels=c(0,1), labels=c("No CHD", "CHD"))
nmes.data$mscd<-factor(ifelse(nmes.data$lc5 == "LC" | nmes.data$chd5 == "CHD", "MSCD", "No MSCD"))
```

First, we want to pick the covariate(s) that we want to include in our propensity score. Here, we will use `age`, but you may want to include different variables for your scores.

We start by fitting a logistic regression model with the main predictor of interest, `eversmk` as the response variable.

```{r}
propModel<-glm(eversmk ~ age, data=nmes.data, family=binomial(link="logit"))
```

Now we want to use this model fit to calculate, for each person in the data set, the predicted probability that they are a smoker, given their age. Think about how you would do this using the model output, given that we just fit the model:
E(log odds of smoking) = beta0 + beta1*age

You could use the estimated coefficients beta0 and beta1, plus the values of the `age` variable, to calculate a predicted log odds of smoking for each person, and then convert this predicted log odds to a probability.

```{r}
```

Fortunately, there are some built-in functions in `R` that make this easier.
```{r}
predLogOdds<-predict(propModel)
predProb<-exp(predLogOdds)/(1+exp(predLogOdds))
```

Now that we have the predicted probabilities, we want to group people according to some quantiles of these probabilities. Here, we will work with quartiles (grouping people into 4 approximately equal-sized groups) but you may want to use something else, like quintiles, on your assignment. First, we need to find the values that separate the quartiles. We can use the `quantile` function for this:
```{r}
# how would you change the code below to make five groups instead of four?
probQuartileCutoffs<-quantile(predProb, probs=c(0,0.25, 0.5, 0.75, 1))
```

Then, we will use the function `cut` to assign each person to a quartile according to their predicted probability.
```{r}
# the code below does not quite work: what is going wrong so that some people are getting assigned na values?
probQuartiles<-cut(predProb, probQuartileCutoffs)
```

Now that we have our quartiles, we can fit our model to explore the relationship between `mscd` and `eversmk` using the propensity scores:
```{r}
glmOut<-glm(mscd ~ eversmk + probQuartiles, data=nmes.data, family=binomial(link="logit"))

```

##### **Part 2: Manipulating tables and basic data reshaping**


```{r}

## here is an example of how you can summarize a factor to get totals for a single variable 
nmes.data %>% group_by(eversmk) %>% summarise(n=n()) 
nmes.data %>% group_by(female) %>% summarise(n=n())

## you can do the same thing but group by two different factors and get counts for each possible combination
nmes.data %>% group_by(female, eversmk) %>% summarise(n=n())

## here is an example of creating a new column using the mutate function; it works group-wise
nmes.data %>% group_by(female, eversmk) %>% summarise(n=n()) %>% mutate(frac = n/sum(n))
nmes.data %>% group_by(eversmk, female) %>% summarise(n=n()) %>% mutate(frac = n/sum(n))
```
This pipe operator (%>%) can be combined with other functions as well, to make nice data visualizations.

```{r}
library(ggplot2)
nmes.data %>% group_by(female, eversmk) %>% summarise(n=n()) %>%
  ggplot(aes(x=eversmk, y=n, fill=female)) + 
  geom_bar(stat="identity")

## use the mutate function to calculate fractions within sex
nmes.data %>% group_by(female, eversmk) %>% summarise(n=n()) %>% mutate(frac = n/sum(n)) %>%
  ggplot(aes(x=eversmk, y=frac, fill=female)) + 
  geom_bar(stat="identity")

## use the mutate function to calculate fractions within smoking (note the different order of the group_by variables)
nmes.data %>% group_by(eversmk, female) %>% summarise(n=n()) %>% mutate(frac = n/sum(n)) %>%
  ggplot(aes(x=eversmk, y=frac, fill=female)) + 
  geom_bar(stat="identity")

## incorporate both chance of disease, sex and smoking status
nmes.data %>% group_by(eversmk, female, lc5) %>% summarise(n=n()) %>% mutate(frac = n/sum(n)) %>%
  ggplot(aes(x=eversmk, y=frac, fill=lc5)) + 
  facet_wrap(~ female) + 
  geom_bar(stat="identity")


## Exercise: see http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization and figure out how to make the bars next to one another


```

How can we use 'mutate' to create a new column that indicates whether they have either lung cancer or CHD?

```{r}
nmes.data<-nmes.data %>% mutate(disease = factor(lc5 == "LC" | chd5 == "CHD", levels=c(FALSE,TRUE), labels=c("No diesease", "Yes disease")))

```

